---
title: "Micro Data Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{md_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pipster)
```
## Overview

This vignette shows an overview of the `pipster` package functions for micro data. Microdata consist of detailed records of individual welfare measures, such as consumption, expenditure, or income, where each observation corresponds to a unique individual, and is accompanied by a sample weight that represents the individual's proportion in the overall population.
`pipster` provides a series of functions to estimate poverty and inequality measures using microdata:

-   `pipmd_pov_headcount()` (FGT0)

-   `pipmd_pov_gap()` (FGT1)

-   `pipmd_pov_severity()` (FGT2)

-   `pipmd_gini()`

-   `pipmd_mld()`

-   `pipmd_watts()`

It also provides a series of functions to calculate distributional measures and to select and validate the best Lorenz curve for subsequent estimation:

-   `pipmd_welfare_share_at()`

-   `pipmd_quantile_welfare_share()`

-   `pipmd_quantile()`

-   `pipmd_validate_lorenz()`

-   `pipmd_select_lorenz()`

## Micro Data Sample

In this vignette, we will explore several typical scenarios in which the pipster package can be effectively utilized. In each of these scenario, we will use a sample dataset with 1000 observations, `pip_md`, available with this package.  The variables are the following:

-   **welfare**: welfare (income or consumption)
-   **weight**: population weights.

## Case 1: Simple Welfare Analysis and Lorenz Curve

### 1.1 Welfare share at a given population share
