---
title: "Micro Data Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{md_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pipster)
```
## Overview

This vignette shows an overview of the `pipster` package functions for micro data. Microdata consist of detailed records of individual welfare measures, such as consumption, expenditure, or income, where each observation corresponds to a unique individual, and is accompanied by a sample weight that represents the individual's proportion in the overall population.
`pipster` provides a series of functions to estimate poverty and inequality measures using microdata:

-   `pipmd_pov_headcount()` (FGT0)

-   `pipmd_pov_gap()` (FGT1)

-   `pipmd_pov_severity()` (FGT2)

-   `pipmd_gini()`

-   `pipmd_mld()`

-   `pipmd_watts()`

It also provides a series of functions to calculate distributional measures and to select and validate the best Lorenz curve for subsequent estimation:

-   `pipmd_welfare_share_at()`

-   `pipmd_quantile_welfare_share()`

-   `pipmd_quantile()`

-   `pipmd_validate_lorenz()`

-   `pipmd_select_lorenz()`

## Micro Data Sample

In this vignette, we will explore several typical scenarios in which the pipster package can be effectively utilized. In each of these scenario, we will use a sample dataset with 1000 observations, `pip_md`, available with this package.  The variables are the following:

-   **welfare**: welfare (income or consumption)
-   **weight**: population weights.

Here is a preview of the data:

```{r data, echo=FALSE}
pip_md |>
  head() |>
  print()
```

## Case 1: Poverty Profiling
`pipster` allows the user to estimate poverty measures quickly and accurately. To demonstrate its use, we can manually calculate FGT(0), FGT(1), and FGT(2), and then replicate it using only `pipster` functions. The Foster-Greer-Thorbecke indices are a family of poverty metrics which can be derived by substituting different values of the parameter $\alpha$ into the following equation:

$$F G T_\alpha=\frac{1}{N} \sum_{i=1}^H\left(\frac{z-y_i}{z}\right)^\alpha$$
where $z$ is the povety line, $N$ the total population, and $H$ the number of poor individuals (with income, or $y_i <= z$).

### 1.1 Poverty Headcount
The poverty headcount, or FGT(1) can be calculated as follows:
$$F G T_0=\frac{1}{N} \sum_{i=1}^H\left(\frac{z-y_i}{z}\right)^0 = \frac{H}{N}$$


```{r headcount-manual}
z = 1.4 # set the poverty line
N = sum(pip_md$weight)
H = sum(pip_md$weight[pip_md$welfare <= z])

FGT0 = H/N

print(paste0("The poverty headcount index is ", round(FGT0*100,2), "%"))
```
In `pipster`, we can simply use the `pipmd_pov_headcount()` function:

```{r headcount-pipster}
pip_FGT0 <- pipmd_pov_headcount(welfare = pip_md$welfare,
                                weight = pip_md$weight,
                                povline = z)

print(paste0("The poverty headcount index is ", round(pip_FGT0$pov_headcount*100,2), "%"))
```

### 1.2 Poverty Gap
The poverty gap, or FGT(1), can be calculated as follows:
$$F G T_1=\frac{1}{N} \sum_{i=1}^H\left(\frac{z-y_i}{z}\right)$$


```{r gap-manual}
# Calculate the shortfall: the distance between the poverty line and the income of the poor, for each poor.
shortfall <- sum((z - pip_md$welfare[pip_md$welfare <= z]) * pip_md$weight[pip_md$welfare <= z])/z

FGT1 <- (1/N)*(shortfall)

print(paste0("The poverty gap index is ", round(FGT1*100,2), "%"))
```


In `pipster`, we can simply use the `pipmd_pov_gap()` function:
```{r gap-pipster}
pip_FGT1 <- pipmd_pov_gap(welfare = pip_md$welfare,
                          weight = pip_md$weight,
                          povline = z)

print(paste0("The poverty gap index is ", round(pip_FGT1$pov_gap*100,2), "%"))
```

### 1.3 Poverty Severity
The poverty severity, or FGT(2), can be calculated as follows:
$$F G T_2=\frac{1}{N} \sum_{i=1}^H\left(\frac{z-y_i}{z}\right)^2$$

```{r severity-manual}
shortfall_squared <- sum(((z - pip_md$welfare[pip_md$welfare <= z]) / z)^2 * pip_md$weight[pip_md$welfare <= z])

FGT2 <- (1/N)*shortfall_squared

print(paste0("The poverty severity index is ", round(FGT2*100,2), "%"))
```

In `pipster`, we can simply use the `pipmd_pov_severity()` function:
```{r severity-pipster}
pip_FGT2 <- pipmd_pov_severity(welfare = pip_md$welfare,
                               weight = pip_md$weight,
                               povline = z)


print(paste0("The poverty severity index is ", round(pip_FGT2$pov_severity*100,2), "%"))
```
